\chapter*{Preface}

\noindent
This workbook is designed to support a one-semester course in \emph{Calculus for Business and Economics}. It is intended for undergraduate students pursuing degrees in business, economics, and related fields who require a strong conceptual and applied understanding of calculus without unnecessary abstraction.

\medskip

\noindent
The primary goal of this workbook is to reinforce key calculus concepts through clear explanations, carefully worked examples, and structured practice problems grounded in business and economic contexts. Emphasis is placed on interpretation, modeling, and problem-solving rather than formal proofs. Topics are selected to align with standard business calculus curricula and to support informed decision-making using mathematical tools.

\medskip

\noindent
This workbook assumes a basic background in algebra and functions. A brief review chapter is included to refresh essential prerequisite skills. Trigonometric functions are intentionally excluded, as they are not central to most business calculus applications. Throughout the workbook, students are encouraged to think critically about the meaning of results and their practical implications.

\medskip

\noindent
Each section includes learning objectives, key definitions and concepts, fully explained solved examples, and practice problems with space provided for student work. Selected answers are collected in an appendix to allow students to check their understanding while still encouraging independent problem-solving.

\medskip

\noindent
This workbook is an Open Educational Resource (OER) created to reduce the cost of instructional materials and increase access to high-quality learning resources. It may be freely used, adapted, and shared in accordance with its open license. Instructors are encouraged to modify and supplement the material to best meet the needs of their students.

\medskip

\noindent
We hope this workbook serves as a useful and accessible resource that supports student learning and confidence in applying calculus to real-world business and economic problems.
